---
title: "Markowitz Model - Optimal Allocation"
author:
-  Juan Carlos Martínez-Ovando
-  jc-mo@tec.mx
date: "Febrero-Junio 2020"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: yes
    self_contained: yes
    theme: cerulean
    highlight: textmate
fig_align: "center"
fig_width: 18
---

# Objetives  {.tabset .tabset-fade .tabset-pills}

* In this note, we shall learn to implement and interpret the optimal allocation problem under Markowitz model perspective, using `R`

# Portfolio allocation problem   {.tabset .tabset-fade .tabset-pills}

## Components

Let us recall that the `portfolio allocation problem` for $n$ assets can be addressed as a **decision theory** problem under *uncertainty* using the following components:

a. **Decision space.-** Defined as the collection of possible portfolio allocation strategies, given by the collection of different proportions available to us to distribute the 100% investment capital amont the $n$ assets, i.e.
$$
\mathcal{\Omega} = \left\{(\omega_1,\ldots,\omega_n): \omega_i \geq 0 \text{ and } \sum_{i=1}^{n}\omega_i=1 \right\}.
$$

> Note that if $(r_1,\ldots,r_n)$ denote the return values of the $n$ assets at a given time, the **return value of the portfolio** would be defined as
$$
\sum_{i=1}^{n}\omega_i r_i = \boldsymbol{\omega}'\boldsymbol{r}.
$$

b. **Uncertainty space.-** In this particular case, since the allocation problem takes place at the begining of time $t$ (a.k.a. towards the end of time $(t-1)$), and the investment takes place along the period $(t-1,t]$, the result of the investment at the end of $t$ is unknown to us at the begining of the same period (begining of $t$ == end of $(t-1)$). 

Thus, what best represent our state of uncertainty corresponds to the actual value of the **return performance** for the $n$ assets along $t$, i.e. the uncertainty space would correspond to the collection of all the return results for the $n$ assets during time $t$, i.e.
$$
\mathcal{R}_t=\left\{ (r_{1t},\ldots,r_{nt}): r_{it} \in \mathbb{R} \right\},
$$
since each one of the assest may or may not have positive returns.

c. **Probability function.-** With the aim of getting a better understanding about the possible outcomes that the $n$ assets may have during time $t$, we make use of the *multivariate Gaussian distribution*. This distribution allow us to understand three key aspects of the possible realization of the $n$ returns:
  - Expected return values, 
  $$
  \hat{\mu}_t=(\mu_{1t},\ldots,\mu_{nt}),
  $$
  where $\mu_{it}=\mathbb{E}(r_{it})$. 
  - Volatility/risk/dispersion of returns,
  $$
  \hat{\sigma}^2_t=(\sigma^2_{1t},\ldots,\sigma^2_{nt}),
  $$
  where $\sigma^2_{it}=var(r_{it})$. 
  - Comovement/synchronicity/covariance among each pair of return assets,
  $$
  \hat{\sigma}_{t}=(\sigma_{ijt}),
  $$
  where $\sigma_{ijt}=cov(r_{it},r_{jt})$. 
  
> The above three characteristics are learned from past data observed before time $t$, i.e. the above characteristics correspond to the point estimates of mean, variance and covariances of the Gaussian distribution using observed data before time $t$.

Thus, we can say that at the begining of time $t$, we can foresee that the possible outcomes of the $n$ return assets can be described as
$$
(r_{it},\ldots,r_{nt}) \sim N\left(\hat{\mu}_t,\hat{\Sigma}_t\right),
$$
where 
$$
\hat{\Sigma}_t = \text{estimated covariance matrix for time }t.
$$

> This point estimates are the ones that you estimated in `Actividad 1` for this subject.

d. **Preference function.-** According to *Markowtiz approach to the portfolio allocation problem*, we can assess the preference of *uncertain return outcomes*, $\boldsymbol{r}_t=(r_{1t},\ldots,r_{nt})$, for a given *decision/allocation strategy*, $\boldsymbol{\omega}=(\omega_{1t},\ldots,\omega_{nt})$, using the utility function

$$
U(\boldsymbol{r}_t,\boldsymbol{\omega}) =
\boldsymbol{r}_t'\boldsymbol{\omega} 
+ \lambda
var(\boldsymbol{r}_t'\boldsymbol{\omega})
$$
where $\lambda$ is a **risk aversion factor** (the greater the $\lambda$ the more affraid to risk we are).

e. **Decision rule.-** This component close with the specification of the decision theory problem associated with the *portfolio allocation problem*. 

> This component is crucial, as it guide us towards the most effective election of the decision variables.

The **decision rule** is demostrated to be **optimal** if we base the election of $(\omega_{1t},\ldots,\omega_{nt})$ with the aim at either

  - **Maximizing the expected gain** associated with $U(\boldsymbol{r}_t,\boldsymbol{\omega})$
  
  or
  
  - **Minimizing the expected loss** associated with $U(\boldsymbol{r}_t,\boldsymbol{\omega})$.
  
> The solution, in this context is referred as  **maximizing expected utility** or **minimizing expected loss**.

In the context of the *portfolio allocation problem*, the aforementioned decision rule is typically referred to the **quadratic optimization problem** given by

\begin{eqnarray}
\text{min}_{\boldsymbol{\omega}\in\Omega} \boldsymbol{\omega}'\hat{\Sigma}\boldsymbol{\omega} && \\
\text{s.t.} && \boldsymbol{\omega}'\boldsymbol{\mu}_t = \mu^*\\
            && \boldsymbol{\omega}'\boldsymbol{1} = 1,\\
\end{eqnarray}
where $\mu^*$ corresponds to a target portfolio return.
  

> With this approach, we select the *portfolio allocation stratgy* based on *minimizing risk* while *achieving a target expected return* in the portfolio.

> In the next example we shall revisit the case of `Petrobras` and `SP500` time series, and its implementation using the package `quadprog` in `R`.

## Example

### Data

Revisiting the exmaple of previous sessions, for `Petrobras` and `SP500` series:

```{r}
library("quantmod")

# Petrobrass series

pbr <- getSymbols("PBR", 
                  src = "yahoo", 
                  from = "2013-01-01", 
                  to = "2017-06-01", 
                  auto.assign = FALSE)
summary(pbr)
plot(pbr[,6])

pbr_ret <- diff(log(pbr[,6]))
pbr_ret <- pbr_ret[-1,]
plot(pbr_ret)

# Basdaq series (USD)
spy <- getSymbols("SPY", 
                  src = "yahoo", 
                  from = "2013-01-01", 
                  to = "2017-06-01", 
                  auto.assign = FALSE)
summary(spy)
plot(spy[,6])

spy_ret <- diff(log(spy[,6]))
spy_ret <- spy_ret[-1,]
plot(spy_ret)

# Merging returns series

port_ret <- cbind(pbr_ret,spy_ret)

```

### Point estimates

Point estimate for the **mean return**, $\hat{\mu}_t$:

```{r}
# Point estimatores
mu.est <- colMeans(port_ret)
mu.est
```

Point estimate for the **covariance matrix of returns**, $\hat{\Sigma}_t$:

```{r}
sigma.est <- cov(port_ret)
sigma.est
```

### Allocation strategy

The solution of the **decision theory problem** is obtained using the `quadprog` package in `R`:

```{r}
if(!require("quadprog")){install.packages("quadprog")}
library("quadprog")
```

Now we set the *limits constraint vector*, considering that we shall base the portfolio allocation strategy with the aim of 

```{r}
symbols = c('SPY','IWM','EFA','EEM','AGG','IYR','GLD')
symbol.names = c('S&P 500','Russell 2000','Europe, Australasia, Far East developed', 'Emerging Markets','Agg bond','REIT','Gold')
getSymbols(symbols, from = '2003-01-01', auto.assign = TRUE)
```

```{r}
candleChart(SPY,theme='white', type='candles')
```

### Nominal and real returns
```{r include=FALSE}
SP500 = c(0.4381,-0.083,-0.2512,-0.4384,-0.0864,0.4998,-0.0119,0.4674,0.3194,-0.3534,0.2928,
              -0.011,-0.1067,-0.1277,0.1917,0.2506,0.1903,0.3582,-0.0843,0.052,0.057,0.183,0.3081,0.2368,
              0.1815,-0.0121,0.5256,0.326,0.0744,-0.1046,0.4372,0.1206,0.0034,0.2664,-0.0881,0.2261,0.1642,
              0.124,-0.0997,0.238,0.1081,-0.0824,0.0356,0.1422,0.1876,-0.1431,-0.259,0.370,0.2383,-0.0698,
              0.0651,0.1852,0.3174,-0.047,0.2042,0.2234,0.0615,0.3124,0.1849,0.0581,0.1654,0.3148,-0.0306,
              0.3023,0.0749,0.0997,0.0133,0.372,0.2382,0.3186,0.2834,0.2089,-0.0903,-0.1185,-0.2197,0.2836,
              0.1074,0.0483,0.1561,0.0548,-0.3658,0.2592,0.148600)
BILLS = c(0.0308,0.0316,0.0455,0.0231,0.0107,0.0096,0.0032,0.0018,0.0017,0.003,0.0008,0.0004,
              0.0003,0.0008,0.0034,0.0038,0.0038,0.0038,0.0038,0.0057,0.0102,0.011,0.0117,0.0148,0.0167,
              0.0189,0.0096,0.0166,0.0256,0.0323,0.0178,0.0326,0.0305,0.0227,0.0278,0.0311,0.0351,0.039,
              0.0484,0.0433,0.0526,0.0656,0.0669,0.0454,0.0395,0.0673,0.0778,0.0599,0.0497,0.0513,0.0693,
              0.0994,0.1122,0.143,0.1101,0.0845,0.0961,0.0749,0.0604,0.0572,0.0645,0.0811,0.0755,0.0561,
              0.0341,0.0298,0.0399,0.0552,0.0502,0.0505,0.0473,0.0451,0.0576,0.0367,0.0166,0.0103,0.0123,
              0.0301,0.0468,0.0464,0.0159,0.0014,0.001300)
BONDS=c(0.0084,0.042,0.0454,-0.0256,0.0879,0.0186,0.0796,0.0447,0.0502,0.0138,0.0421,0.0441,
            0.054,-0.0202,0.0229,0.0249,0.0258,0.038,0.0313,0.0092,0.0195,0.0466,0.0043,-0.003,0.0227,
            0.0414,0.0329,-0.0134,-0.0226,0.068,-0.021,-0.0265,0.1164,0.0206,0.0569,0.0168,0.0373,0.0072,
            0.0291,-0.0158,0.0327,-0.0501,0.1675,0.0979,0.0282,0.0366,0.0199,0.0361,0.1598,0.0129,-0.0078,
            0.0067,-0.0299,0.082,0.3281,0.032,0.1373,0.2571,0.2428,-0.0496,0.0822,0.1769,0.0624,0.150,
            0.0936,0.1421,-0.0804,0.2348,0.0143,0.0994,0.1492,-0.0825,0.1666,0.0557,0.1512,0.0038,0.0449,
            0.0287,0.0196,0.1021,0.201,-0.1112,0.084600)
CPI=c(-0.0115607,0.005848,-0.0639535,-0.0931677,-0.1027397,0.0076336,0.0151515,0.0298507,
          0.0144928,0.0285714,-0.0277778,0.0000,0.0071429,0.0992908,0.0903226,0.0295858,0.0229885,
          0.0224719,0.1813187,0.0883721,0.0299145,-0.0207469,0.059322,0.0600,0.0075472,0.0074906,
          -0.0074349,0.0037453,0.0298507,0.0289855,0.0176056,0.017301,0.0136054,0.0067114,0.0133333,
          0.0164474,0.0097087,0.0192308,0.0345912,0.0303951,0.0471976,0.0619718,0.0557029,0.0326633,
          0.0340633,0.0870588,0.1233766,0.0693642,0.0486486,0.0670103,0.0901771,0.1329394,0.125163,
          0.0892236,0.0382979,0.0379098,0.0394867,0.0379867,0.010979,0.0443439,0.0441941,0.046473,
          0.0610626,0.0306428,0.0290065,0.0274841,0.026749,0.0253841,0.0332248,0.017024,0.016119,
          0.0268456,0.0338681,0.0155172,0.0237691,0.0187949,0.0325556,0.0341566,0.0254065,0.0408127,
          0.0009141,0.0272133,0.0149572)
GOLD = c(0.000000000,0.000000000,0.000000000,0.000000000,0.000000000,0.447002860,0.079661828,
             0.000000000,0.000000000,0.000000000,0.000000000,0.000000000,-0.014388737,0.028573372,0.000000000,
             0.027779564,-0.006872879,0.027212564,0.026491615,0.117056555,-0.023530497,-0.036367644,
             -0.006191970,-0.006230550,-0.033039854,-0.086306904,-0.007067167,-0.002840911,0.001421464,
             0.001419447,0.000000000,0.000000000,0.034846731,-0.027779564,-0.004234304,-0.002832863,
             0.002832863,0.004234304,-0.002820876,0.002820876,0.203228242,-0.059188871,-0.052577816,
             0.136739608,0.358646094,0.511577221,0.545727802,-0.132280611,-0.253090628,0.168898536,
             0.265477915,0.464157559,0.689884535,-0.285505793,-0.201637346,0.120144312,-0.163629424,
             -0.127202258,0.149181164,0.192236014,-0.020385757,-0.134512586,0.005221944,-0.058998341,
             -0.051002554,0.045462374,0.064538521,0.002600782,0.010336009,-0.155436854,-0.121167134,
             -0.052185753,0.000000000,-0.028987537,0.133990846,0.157360955,0.119003292,0.084161792,
             0.308209839,0.142551544,0.220855221,0.110501915,0.228258652)
fnominal=data.frame(stocks=SP500, bills=BILLS, bonds=BONDS, gold=GOLD, CPI=CPI)
freal=data.frame(stocks=SP500-CPI, bills=BILLS-CPI, bonds=BONDS-CPI, gold=GOLD-CPI)
```

```{r}
realreturns = apply(freal, 2, mean)
realreturnspct = realreturns*100
realreturnspct
```

### Real volatility (covariance)

```{r}
realsds = apply(freal, 2, sd)
realsdspct = realsds*100
realsdspct
```

```{r include=FALSE}
ia = list()
ia$n = length(freal)
ia$annual.factor = 1
ia$symbols = names(freal)
ia$symbol.names = names(freal)
ia$hist.returns = freal
ia$arithmetic.return = apply(ia$hist.returns, 2, mean, na.rm = T)
ia$arithmetic.return = (1 + ia$arithmetic.return)^ia$annual.factor - 1
ia$geometric.return = apply(ia$hist.returns, 2, function(x) prod(1+x)^(1/length(x))-1 )
ia$geometric.return = (1 + ia$geometric.return)^ia$annual.factor - 1
ia$risk = apply(ia$hist.returns, 2, sd, na.rm = T)
ia$risk = sqrt(ia$annual.factor) * ia$risk
ia$correlation = cor(ia$hist.returns, use = 'complete.obs', method = 'pearson')
ia$cov = cov(ia$hist.returns, use = 'complete.obs', method = 'pearson')
ia$expected.return = ia$arithmetic.return


con = gzcon(url('http://www.systematicportfolio.com/sit.gz', 'rb'))
source(con)
close(con)

n = ia$n

constraints = new.constraints(n, lb = 0, ub = 1)

constraints = add.constraints(rep(1, n), 1, type = '=', constraints)
ef.risk = portopt(ia, constraints, 50, 'Historical', equally.spaced.risk = T)
```

### Optimal solution

```{r}
layout( matrix(1:2, nrow = 2) )
plot.ef(ia, list(ef.risk), portfolio.risk, T, T)
```